---
title:
  formatted: "An \\proglang{R} Markdown Template for JSS Submissions"
  plain: "An R Markdown Template for JSS Submissions"
author: 
  - name: Martin Lysy
    affiliation: University of Waterloo
    address:
      - 200 University Avenue West
      - Ontario, Canada
    email: \email{mlysy@uwaterloo.ca}
  - name: First Student
    affiliation: University of Waterloo
  - name: Second Student
    affiliation: University of Waterloo
abstract: >
  Here is the abstract.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
date: "`r Sys.Date()`"
documentclass: jss
classoption: article
output: 
  bookdown::pdf_document2:
    toc: false
    template: jss-template.tex
    includes:
      in_header: jss-includes.tex
    citation_package: natbib
    # keep_tex: true
    highlight: tango
---

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(comment="",
                      prompt = TRUE,
                      R.options = list(prompt = "R> ",
                                       continue = "+  ",
                                       width = 70,
                                       useFancyQuotes = FALSE))

#' Embed arbitary code files and typset them correctly. 
embed_file <- function(file, lang) {
  cat(paste0("```", lang), readLines(file), "```", sep = "\n")
}
```

# Numbering {#sec:arbitrary}

## Equations

Here's a reference to equation \@ref(eq:xyz).
\begin{equation}
x + y = z.
(\#eq:xyz)
\end{equation}
Here's an unnumbered equation:
\[
a/b = c.
\]
Here's some inline math: $y = \exp(x^2 - 1)$.

## Figures

Reference to Figure \@ref(fig:plot).  Also an example of embedding \proglang{R} code.
```{r plot, fig.cap = "A simple plot."}
x <- {function(y) {
  y + 1:10
}}(3)
x
plot(x, pch = x, col = x)
```

## Tables

Reference to Table \@ref(tab:mtcars).
```{r mtcars, echo = FALSE}
knitr::kable(head(mtcars[, 1:8], 5), booktabs = TRUE,
             caption = "A table of the first 5 rows of the mtcars data.")
```

## Sections

Can use built-in labels, as in Section \@ref(tables), or custom labels, as in Section \@ref(sec:arbitrary).  The latter is highly recommended.

# Math in Headings

**Solved**.  Here's an example:

\subsection[C++ File alpha + beta]{\proglang{C++} File $\alpha + \beta$}

Below describes the old problem and workaround.

- LaTeX in heading doesn't seem to work.  [Here's](https://github.com/jgm/pandoc/issues/3555) how far I got with this.
- In fact, seems to be a pure LaTeX problem, in that `latexmk` compile of tex output takes several tries to get it right.
- Working solution: Use `\subsection[plaintext]{latex}` instead of Markdown section.
- ~~New problem: `\tightlist` not defined...~~

# Citations

The work of @Zeileis+Kleiber+Jackman:2008 has been cited multiple times [@Chambers+Hastie:1992].

# Embedded Files

## C++ File

```{r, echo = FALSE, results = "asis"}
embed_file("foo.h", "cpp")
```

## R Markdown File

```{r, echo = FALSE, results = "asis"}
embed_file("foo.Rmd", "bash")
```

<!-- references -->

\bibliography{refs}

<!-- appendix -->

# (APPENDIX) Appendix {-}

# More information

This will be Appendix A.

# One more thing

This will be Appendix B.
